Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-02-14 ms.Rnw
start examples; more edits

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(viridis)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.10.2")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

@@

Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
We will frequently need to refer to the left-limit of such a process.
Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
\begin{equation*}
  \Zrt_t\colonequals
  \begin{cases}
    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
    \Zr_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\Zrt_t$ is thus left-continuous with right limits.

If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.

@@

Motivating examples: compartmental models.
Wide variety of models.
Linear chain trick.
Migration, superspreading, competition between strains.

@@

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}
  \label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.

@@

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}
  \label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.

@@

In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.

For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.

@@

In the following, it will be useful to break the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

@@

@@

\subsubsection{SIIR model}

@@

For our purposes, a \emph{genealogy}, $\G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
A genealogy includes three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $\G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{\G}{T}$ denote the set of all internal and sample node-times of the genealogy $\G_T$;
we refer to these as \emph{genealogical event times}.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $\G_T$, to be an ordered pair $(\y,\z)$.
The $\z$ element specifies the genealogy's tree structure, as follows.
Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
Partition \emph{fineness} defines a partial order on $\func{Z}$.
Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
The tree structure of $\G_T$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
An element $b\in\z_t$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.

The $\y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
We will also need to count the number of nodes along each lineage:
let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=t(a)$ to $t=0$.
Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
Formally, we define the coloring component $\y$ as follows.
Let $\Yspace=\Demes\times\Zp$.
For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
we say that $\y$ is \emph{compatible} with $\z$ if,
for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
$\y$ is \cadlag\ and,
whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
we have $y^m_{t_1}\le{y^m_{t_2}}$.
Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$.
\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}

@@

We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a related construction that is equivalent to the one presented here.

@@

Application of these rules at each jump of $\Xr_t$ constructs a chain of genealogies $\Grh_k$.
In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $(\Trh_k,\Xrh_k)$ for all $k$.
Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
We refer to this property as the \emph{exchangeability} of lineages within a deme.

@@

\begin{figure}
  <<upo,fig.dim=c(6,8),out.width="60%",results="hide">>=
  freeze(
    seed=522390503,
    simulate(
      "SEIR",
      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
      S0=10,E0=1,I0=1,R0=0,
      time=10
    )
  ) -> x

  pal <- c("#00274CFF","#FFCB05FF")

  plot_grid(
    A=x |>
      plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
      lims(x=c(NA,10),y=c(3,NA))+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    B=x |>
      plot(points=TRUE,prune=TRUE,obscure=FALSE,palette=pal)+
      lims(x=c(NA,10),y=c(3,NA))+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    C=x |>
      plot(points=TRUE,prune=TRUE,obscure=TRUE,palette="#B3B3B3FF")+
      lims(x=c(NA,10),y=c(3,NA))+
      geom_vline(xintercept=10,linewidth=0.2,color="black")+
      labs(x="time"),
    ncol=1,
    align="hv",axis="tblr",
    labels="AUTO"
  )
  @
  \caption{
    Unpruned, pruned, and obscured genealogies from a single realization of the genealogy process induced by the SEIRS model depicted in \cref{fig:example_models,fig:markov_state}.
    \textbf{(A)} The full genealogy at time $t=10$.
    Tip nodes, corresponding to lineages alive at time $t=10$ are indicated with black points.
    Blue points represent samples;
    green points, internal nodes.
    Branches are colored according to the deme in which the corresponding lineage resided at that point in time:
    blue denotes $\lab{E}$ and yellow, $\lab{I}$.
    \textbf{(B)} The full genealogy is \emph{pruned} by removing all tip nodes and then recursively pruning away childless internal nodes.
    Sample nodes are never removed.
    \textbf{(C)} A genealogy is \emph{obscured} by removing all deme information from lineage histories:
    the colors are erased, as are extraneous internal nodes.
    See the text (\cref{sec:genealogy,sec:pruning,sec:obscuration}) for more detail.
    \label{fig:upo}
  }
\end{figure}

@@

Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
Observe that a pruned genealogy is a colored genealogy:
it retains information about where among the demes each of its lineages was through time.
Note also that a pruned genealogy $\Pr_t$ is characterized by functions $\Pr_t^\y$ and $\Pr_t^\z$ just as an un-pruned genealogy is.
Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t$ is Markov.

>>>>

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(viridis)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.5")
stopifnot(packageVersion("phylopomp")>="0.10.3")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

@@

Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $\Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
In the case of continuous-time processes, we will assume that sample paths are \cadlag\ \ie, right-continuous with left limits.
We will frequently need to refer to the left-limit of such a process.
Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
\begin{equation*}
  \Zrt_t\colonequals
  \begin{cases}
    \displaystyle\lim_{s\,\uparrow\,{t}}\;\Zr_{s}, & t>0,\\[2ex]
    \Zr_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\Zrt_t$ is thus left-continuous with right limits.

If $\Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its jump times in the interval $[0,t]$.

@@

Motivating examples:
compartmental models in ID epidemiology,
phylogenetics and systematics.
Wide variety of models are of interest (\cref{fig:examples}).
Linear chain trick.
Migration, superspreading, competition between strains.

@@

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state density, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}
  \label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
\AAK{[We could mention that we use the continuity of $\alpha$ with respect to $t$ in some of the following, but that this condition can be relaxed.]}
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\Xr_t$ is non-explosive, so that $\Prob{\Kr_t<\infty}=1$ for all $t$.

@@

The above may be compactly summarized by stating that if $v(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}
  \label{eq:kfe}
  \frac{\partial{v}}{\partial{t}}(t,x)=\int\!v(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!v(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $v(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}\!{v(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.

@@

In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
In particular, we suppose that there are a countable set of subpopulations that may differ in their vital rates, but within each of which, individual lineages are statistically identical.
We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
We define the \emph{deme occupancy} function $n:\Demes\times\Xspace\to\Zp$ so that
for $i\in\Demes$, $x\in\Xspace$, $n_i(x)$ is the number of lineages in deme $i$ when the population is in state $x$.

@@

In the following, it will be useful to break the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
\Cref{fig:markov_state} shows an example with five distinct marks.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

\subsubsection{SIRS model}

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
Its state vector is $x=(S,I,R)$ and its KFE is
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,I-1,R)
    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,I,R)\\
    &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
    -\gamma\,I\,v(t,S,I,R)\\
    &+\omega\,(R+1)\,v(t,S-1,I,R+1)
    -\omega\,R\,v(t,S,I,R).
  \end{split}
\end{equation*}
Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,E-1,I,R)
    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,E,I,R)\\
    &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
    -\sigma\,E\,v(t,S,E,I,R)\\
    &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
    -\gamma\,I\,v(t,S,E,I,R)\\
    &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
    -\omega\,R\,v(t,S,E,I,R),
  \end{split}
\end{equation*}
where $N=S+E+I+R$ is the total population size.
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

\subsubsection{Two-strain competition model}

A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
This model has KFE
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}=
    &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
    -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
    -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
    -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
    -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
    -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
    -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
    -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
    -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
  \end{split}
\end{equation*}

\begin{equation*}
  \frac{\partial{v}}{\partial{t}}=
  \lambda\,(N-1)\,v(t,N-1)
  -\lambda\,N\,v(t,N)
  +\mu\,(N+1)\,v(t,N+1)
  -\mu\,N\,v(t,N)
\end{equation*}

@@

\begin{figure}
  \begin{center}
    <<geneal0>>=
    freeze(
      seed=522390503,
      simulate(
        "SEIR",
        Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
        S0=10,E0=1,I0=1,R0=0,
        time=10
      )
    ) -> x
    @
    <<geneal1,dependson="geneal0",fig.dim=c(4,2),out.width="50%">>=
    pal <- c("#00274CFF","#FFCB05FF")
    x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
      geom_vline(xintercept=10,linewidth=0.2,color="black")
    @
  \end{center}
  \caption{
    $\time(\G)=10$.
    Two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
    Extant lineages are denoted with black dots;
    blue dots indicate samples.
    \label{fig:geneal}
  }
\end{figure}

A \emph{genealogy}, $\G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(\G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}A).
A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
\item \emph{sample nodes}, which represent labeled samples; and
\item \emph{root nodes}, at the base of each tree.
\end{inparaenum}
Each node $a$ is associated with a specific time, $\time(a)$.
In particular, if $a$ is a tip node in $\G$, then $\time(a)=\time(\G)$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
We let $\event{\G}{}$ denote the set of all internal and sample node-times of the genealogy $\G$;
we refer to these as \emph{genealogical event times}.

Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme (\cref{fig:geneal}A).

Formally, we define a genealogy, $\G$, to be an ordered pair $(\y,\z)$.
The $\z$ element specifies the genealogy's tree structure, as follows.
Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
Partition \emph{fineness} defines a partial order on $\func{Z}$.
Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
If $\time(\G)=T$, the tree structure of $\G$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
An element $b\in\z_t$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.

The $\y$ element of $\G$ specifies the coloring of branches and location of nodes.
In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
We will also need to count the number of nodes along each lineage:
let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=\time(a)$ to $t=0$.
Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
Formally, we define the coloring component $\y$ as follows.
Let $\Yspace=\Demes\times\Zp$.
For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
we say that $\y$ is \emph{compatible} with $\z$ if,
for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
$\y$ is \cadlag\ and,
whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
we have $y^m_{t_1}\le{y^m_{t_2}}$.
Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$ and $\func{Y}_t(\z)\coloneq\CondSet{y_t}{y\in\func{Y}(\z)}$.
\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}

@@

@@

@@

We now show how a given population process induces a stochastic process, $\Gr_t$, on the space of genealogies.
In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a related construction that is equivalent to the one presented here.

@@

Application of these rules at each jump of $\Xr_t$ constructs a chain of genealogies $\Grh_k$.
In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gr_t$.
It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $(\Trh_k,\Xrh_k)$ for all $k$.
Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
We refer to this property as the \emph{exchangeability} of lineages within a deme.

@@

\begin{figure}
  <<geneal2,dependson="geneal0",fig.dim=c(4,6),out.width="50%">>=
  pal <- c("#00274CFF","#FFCB05FF")
  plot_grid(
    A=x |>
      plot(
        points=TRUE,prune=FALSE,obscure=FALSE,
        ladderize=FALSE,palette=pal
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    B=x |>
      plot(
        points=TRUE,prune=TRUE,obscure=FALSE,
        ladderize=FALSE,palette=pal
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    C=x |>
      plot(
        points=TRUE,prune=TRUE,obscure=TRUE,
        ladderize=FALSE,palette="#B3B3B3FF"
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    ncol=1,
    align="hv",axis="tblr",
    labels="AUTO"
  )
  @
  \caption{
    Unpruned, pruned, and obscured genealogies from a single realization of the genealogy process induced by the SEIRS model depicted in \cref{fig:example_models,fig:markov_state}.
    \textbf{(A)} A realization of the unpruned genealogy process $\Gr_t$ is shown at $t=10$.
    Tip nodes, corresponding to lineages alive at time $t=10$ are indicated with black points.
    Blue points represent samples;
    green points, internal nodes.
    Branches are colored according to the deme in which the corresponding lineage resided at that point in time:
    blue denotes $\lab{E}$ and yellow, $\lab{I}$.
    \textbf{(B)} The genealogy is \emph{pruned} by deleting all tip nodes and then recursively pruning away childless internal nodes.
    Sample nodes are never removed.
    \textbf{(C)} A genealogy is \emph{obscured} by effacing all deme information from lineage histories:
    the colors are erased, as are extraneous internal nodes.
    See the text (\cref{sec:genealogy,sec:pruning,sec:obscuration}) for more detail.
    \label{fig:upo}
  }
\end{figure}

@@

Given a genealogy $\G$, one obtains the \emph{pruned genealogy}, $\P=\prune(\G)$ by first dropping every tip node and then recursively dropping every childless internal node (\cref{fig:upo}A--B).
In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
Observe that a pruned genealogy is a colored genealogy:
it retains information about where among the demes each of its lineages was through time (\cref{fig:upo}B).
Note also that a pruned genealogy $\P$ is characterized by functions $\P^\y$ and $\P^\z$ just as an un-pruned genealogy is.
Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t=\prune(\Gr_t)$ is Markov, even though the unpruned genealogy process, $\Gr_t$, is not.
